Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-04 manuscript/climate_sensitivity_draft.Rmd
worked on discussion related to tree height

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006], and this may become a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

@@

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-27% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (*Greenwood et al. 2007- doi: @greenwood_tree_2017; @anderegg_meta-analysis_2016). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height. To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. They also face different microenvironments, which are partially related to canopy position. Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig. 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to be lower in the canopy than in the understory [@koike_leaf_2001], implying greater water costs of $CO_2$ uptake. Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006]. We show that height is a far stronger predictor than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

Beyond microenvironment, multiple tree characteristics covary with height. To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, fine root biomass and rooting volume, sapwood area, and transpiration (e.g., @andersonteixeira_size-related_2015; @chen_allometry_2019). These work in combination to shape whole-plant hydraulic function (**REFS-KAT: check McDowell scaling paper,others?**; @sperry_species-level_2012; @couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees. However, height is arguably the most biophysically meaningful tree-size variable shaping drought response. Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]. Our analysis focuses on species-level comparisons and does not characterize the role of variation with height. On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height. In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**). Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**... (**REFS**). It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).

>>>>

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**). However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], and this may become a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

@@

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature. While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-27% for each individual drought (Table 5). Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*). Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (@greenwood_tree_2017; @anderegg_meta-analysis_2016). Nevertheless, our results are consistent with findings from more extreme droughts.

Our analysis indicates that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response. Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. Vertical gradients in stem and leaf hydraulic traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges. Taller trees also face different microenvironments (Fig. 2a-b), in part because they are more likely to be in dominant canopy positions (Fig. 2d). Even under non-drought conditions, evaporative demand increases with tree height, and tall trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]). Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height. Correlation between height and canopy position (Fig. 2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment. However, signficant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position (Tables 1,4,5). Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4). However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*. The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]). It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**). The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**). Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**... (**REFS**). It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).

Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height. As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits. If trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
